The Spatial and Temporal Domains of Modern Ecology

### 2017-05-25 Rmd/space-time.Rmd
Finalized (or nearly) SI, and main text revised up to final section, and final part of abstract.

# plot(rrs, xlim = c(-5, 15), ylim = c(-5, 15), 
#      xlab = "Sample interval - log10(days)", 
#      ylab = "Plot resolution - log10(m^2)")
# plot(ers, box = FALSE, xlim = c(-5, 15), ylim = c(-5, 15), 
#      xlab = "Study duration - log10(days)", ylab = "Sampled area - log10(ha)")
# dev.off()
```

>>>>

# plot(rrs, xlim = c(-5, 15), ylim = c(-5, 15), 
#      xlab = "Sample interval - log10(days)", 
#      ylab = "Plot resolution - log10(m^2)")
# plot(ers, box = FALSE, xlim = c(-5, 15), ylim = c(-5, 15), 
#      xlab = "Study duration - log10(days)", ylab = "Sampled area - log10(ha)")
# dev.off()

### 2017-05-25 paper/SI.tex
Finalized (or nearly) SI, and main text revised up to final section, and final part of abstract.

\usepackage{scicite}
\usepackage{times}
\usepackage{graphicx}
\usepackage{lineno}
\usepackage{longtable}
%\usepackage{multicolumn}
\usepackage[skip=2pt,font=normalsize]{caption}

@@

@@

\subsubsection*{Calibration and consistency}
\vspace{-10pt}
Most studies did not explicitly report values for all four of the assessed scales, and thus interpretation and judgement had to be applied to develop reasonable estimates for their values. The FAQ provided the protocol we followed, and was initially developed following consultation between observers prior to the commencement of review. We conducted an iterative process of calibration to ensure consistency and reliability of estimates. First, we used the calibration set to calculate between-observer variability with respect to paper selection/rejection and the estimation of scales.  Based on this, the lead author reviewed individual records in each observer's calibration set, flagged values where the estimation procedure departed from the protocol, and returned these to observers for re-estimation, without providing an estimate of the actual value. Instead, the relevant section of the protocol was highlighted, and further explanation and clarifying discussion undertaken as needed. Protocol language was adjusted for clarity during this process, and new items added to cover circumstances that had not been addressed by the initial version. The variability measures (see Results) were re-calculated after each iteration.

@@

\subsubsection*{Accounting for estimation uncertainty}
\vspace{-10pt}
Two major sources of uncertainty affected our estimation of observational scales: 1) unclear documentation of observational scales in the reviewed studies, 2) variation between observers in estimating observational scales. We estimated and accounted for these uncertainties in several ways. First, we estimated the degree of between-observer variability based on each observer's results from reviewing the 20-paper calibration set. We calculated how well observers agreed regarding paper inclusion/exclusion, how many extractable observations there were per included paper, and what the coefficient of variation was across all observers' estimates of scale. We also recorded when observations were reported in a study with an unclear or missing scale value.

We used the between-observer coefficients of variation for each dimension as the basis for randomly perturbing the scale values for each of the 371 observations over 1000 iterations. For each observation at each iteration, we perturbed its scale value in each dimension by randomly selecting a percentage value p that fell between 100 + y and 100 - y, where y was the between-observer coefficient of variation (expressed as a percentage) for a given observational dimension, and multiplying that observation's scale value by that proportion. This perturbed set of observations provided a basis for estimating uncertainty within our extracted set of observational scales.

\subsubsection*{Scale metrics}
\vspace{-10pt}
In presenting results (Fig. 1 and 2 in main text), we log-transformed (base 10) the observational scales within each of the 1000 perturbed ensemble members in order to account for the large range in scale values. We calculated histograms for each observational dimension from each of the log-transformed ensemble members, calculating the mean percentage density estimate for each bin across all 1000 histograms, as well as the upper and lower 2.5th percentile values for each bin (Fig. 1 main text).

To reveal the densities of observations within two dimensions (Fig. 2 main text), we used the splancs package \cite{rowlingson_splancs:_1993} of R \cite{r_development_core_team_r:_2011} to calculate a kernel density estimate of the log-transformed values across all ensemble members, using a bandwidth of 1 onto a 0.1 resolution image to provide a smoothed result that served to more effectively highlight domains in which ecological observations are concentrated. Bandwidths of varying resolutions were tested on kernel density estimates were tested on kernel density estimates of sampling interval versus plot resolution (see Supplementary Results section).

@@

Second, we calculated the intra-class correlation coefficient \cite{bartko_intraclass_1966} to assess agreement between observers regarding the number of ecological observations that could be extracted from each paper (multiple ecological observations were reported in many studies, and we listed observations as separate records if they varied on one or more dimensions).  The coefficient, calculated using the IRR package \cite{gamer_irr:_2012} of R, was 0.70 (F = 15.4, p$<$0.0001, 95\% confidence interval = 0.54 - 0.85)

Finally, we calculated the coefficient of variation (CV) between observers' estimates of scales for each dimension, first across all six observers' mean scale estimates, and then as the average CV among observers' estimates of each individual record (Table S2).

\begin{table}[ht]
\captionsetup{width=0.725\linewidth}
\caption{Estimates of variability calculated from each observer's estimates of the spatial and temporal scales of ecological observations reported within a common set of the 20 papers used for calibration. Variability is expressed as the coefficient of variation (CV; standard deviation divided by mean multiplied by 100) between each observer's overall mean, and as the mean CV of observers' estimates for individual records. }
\vspace{-4 pt}
\begin{center}
\begin{tabular}{rrrrr}
\hline
& \multicolumn{2}{c}{\textbf{Spatial}} & \multicolumn{2}{c}{\textbf{Temporal}}\\\cline{2-5}
Value & Resolution & Extent & Interval & Duration \\ 
\hline
CV of overall mean & 54 & 82 & 54  & 110 \\
Mean of record-wise CV & 58 & 68 & 99 & 124 \\
\hline
\end{tabular}
\end{center}
\label{default}
\end{table}%

@@

@@

\subsubsection*{Trends in observational methods by year}
\vspace{-10pt}

@@

\begin{figure}[!ht]
%\begin{wrapfigure}{c}{1\textwidth}
\center{\includegraphics[width=0.8\textwidth]{figures/dim_by_yr.png}}
\vspace{5 pt}
\caption{Trends in observational scales by year of publication. The coefficient of a weighted (by number of studies in each year) linear regression, fit to the logarithm (base 10) of the mean scale values (calculated for each publication year) for the six assessed dimensions ispresented at the top of each plot, as well as the model coefficient of determination and p-value. }
\label{afoto1}
\end{figure}
%\clearpage

\subsubsection*{Choice of bandwidth in kernel density estimation}
\vspace{-10pt}
Figure S1 indicates the effect that varying bandwidth has on the appearance of the kernel density estimates. The smallest bandwidth (0.4) tested (Fig. S2 top) shows that the primary observational concentrations described in the main text are evident but harder to discern.  Most difficult to discern is the oblong concentration of observations that in Fig. 1A (main text) is bounded on the lower right at monthly to yearly intervals and 100-1000 m$^2$ resolutions and on the upper left by near-daily to monthly observations and 0.1-10 ha resolutions. With the smaller bandwidth this concentration appears as two separate patches (Fig. S2 top left), but with 0.7 bandwidth applied becomes coherent (Fig. S3 middle left).

\begin{figure}[!ht]
%\begin{wrapfigure}{c}{1\textwidth}
\includegraphics[width=0.9\textwidth]{figures/res_v_extent_ksize2.pdf}
\vspace{10 pt}
\caption{Two-dimensional kernel density estimates of observational densities within the domains defined by sampling interval and spatial resolution (left column) and temporal duration and spatial extent (right column), applied to log-transformed values of each observational dimension. Rows indicate the effects of selecting different bandwidths: 0.4 (top row); 0.7 (middle row); 1 (bottom row). }
\label{afoto1}
\end{figure}

>>>>

\usepackage{scicite}
\usepackage{times}
\usepackage{graphicx}
\usepackage{lineno}
\usepackage{longtable}
%\usepackage{multicolumn}
\usepackage[skip=2pt,font=normalsize]{caption}
\usepackage[LGRgreek]{mathastext}
\usepackage{multirow}

@@

\usepackage{array}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

@@

\subsubsection*{Calibration and consistency}
\vspace{-10pt}
Most studies did not explicitly report values for all the assessed scales, and thus interpretation and judgement had to be applied to develop reasonable estimates for their values. The FAQ provided the protocol we followed, and was initially developed following consultation between observers prior to the commencement of review. We conducted an iterative process of calibration to ensure consistency and reliability of estimates. First, we used the calibration set to calculate between-observer variability with respect to paper selection/rejection and the estimation of scales.  Based on this, the lead author reviewed individual records in each observer's calibration set, flagged values where the estimation procedure departed from the protocol, and returned these to observers for re-estimation, without providing an estimate of the actual value. Instead, the relevant section of the protocol was highlighted, and further explanation and clarifying discussion undertaken as needed. Protocol language was adjusted for clarity during this process, and new items added to cover circumstances that had not been addressed by the initial version. The variability measures (see Results) were re-calculated after each iteration.

@@

\subsubsection*{Accounting for estimation uncertainty}
\vspace{-10pt}
Two major sources of uncertainty affected our estimation of observational scales: 1) unclear documentation of observational scales in the reviewed studies, 2) variation between observers in estimating observational scales. We estimated and accounted for these uncertainties in several ways. First, we calculated the degree of between-observer variability based on each observer's results from reviewing the 20-paper calibration set. We calculated how well observers agreed regarding paper inclusion/exclusion, how many extractable observations there were per included paper, and what the coefficient of variation was across all observers' estimates of scale. We also recorded when observations were reported in any study (calibration set or otherwise) with an unclear or missing scale value.

We used the between-observer coefficients of variation for each dimension as the basis for randomly perturbing the scale values for each of the 371 observations over 1000 iterations. For each observation at each iteration, we perturbed its scale value in each dimension by randomly selecting (from a uniform distribution) a percentage value p that fell between 100 + y and 100 - y, where y was the between-observer coefficient of variation (expressed as a percentage) for a given observational dimension, and multiplying that observation's scale value by that proportion. This perturbed set of observations provided a basis for estimating uncertainty within our extracted set of observational scales.

\subsubsection*{Scale metrics}
\vspace{-10pt}
In presenting results (Fig. 1 and 2 in main text), we log-transformed (base 10) the observational scales within each of the 1000 perturbed ensemble members in order to account for the large range in scale values. We calculated histograms for the four primary observational dimensions from each of the log-transformed ensemble members, calculating the mean percentage density estimate for each bin across all 1000 histograms, as well as the upper and lower 2.5th percentile values for each bin (Fig. 1 main text).

To reveal the densities of observations within two dimensions (Fig. 2 main text), we used the splancs package \cite{rowlingson_splancs:_1993} of R \cite{r_development_core_team_r:_2011} to calculate a kernel density estimate of the log-transformed values across all ensemble members, using a bandwidth of 1 on a 0.1 resolution image to provide a smoothed result that served to more effectively highlight domains in which ecological observations are concentrated. Bandwidths of varying resolutions were tested on kernel density estimates were tested on kernel density estimates of sampling interval versus plot resolution (see Supplementary Results section).

To compare the differences between actual extent and extent and actual duration and duration (Fig. 3 main text) we calculated the magnitude of difference (decade) between each pair as:

\begin{equation}
decade = log_{10}(x) - log_{10}(y) = log_{10}\left(\frac{x}{y}\right)
\end{equation}

\noindent Where x is either extent or duration and y is its actual counterpart. We then evaluated how the magnitudes of difference varied with increasing values of actual extent/duration, using box plots to summarize decades within the same bins used for extent and duration (Fig. 1B and D main text). Decades were calculated for each pair for all bootstrap replicates. The whiskers in the summary box plots (Fig. 3 main text) fall slightly below 0 in some bins, even though extent and duration can never be smaller than their actual value. This is an artifact of the perturbation applied to estimated values, which in some instances caused extent/duration values to fall below their actual value. We retained these discrepancies rather than forcing such cases to zero to avoid biasing the summary statistics.

\subsubsection*{Trends in methods and scale}
To assess the degree to which observational methods changed over the course of the 10 years worth of studies we surveyed, we calculated the percentage of observations made using remote sensing, general field methods, and automated \emph{in situ} methods, and fit a linear regression between these percentages and publication year, weighting the regression by the total number of observations in each year.  We performed the same analysis for the four primary dimensions (resolution, extent, interval, and duration) in order to assess whether there were any trends in observational scales.

@@

Second, we calculated the intra-class correlation coefficient \cite{bartko_intraclass_1966} to assess agreement between observers regarding the number of ecological observations that could be extracted from each paper (multiple ecological observations were reported in many studies, and we listed observations as separate records if they varied on one or more dimensions).  The coefficient, calculated using the IRR package \cite{gamer_irr:_2012} of R, was 0.71 (F = 15.4, p$<$0.0001, 95\% confidence interval = 0.54 - 0.85)

Finally, we calculated the coefficient of variation (CV) between observers' estimates of scales for each dimension, first across all observers' mean scale estimates, and then as the average CV among observers' estimates of each individual record (Table S2). CV values were estimated from all six observers for resolution, actual extent, and interval, from five observers (all but Choi) for duration, and from three observers (Elsen, Estes, and Treuer) for extent. The smaller numbers for duration and extent reflect that initial efforts were focused on estimating resolution, interval, and actual extent and actual duration, thus fewer observers were available to assess extent and extent.

@@

\begin{table}[ht]
\captionsetup{width=1\linewidth}
\caption{Estimates of variability calculated from each observer's estimates of the spatial and temporal scales of ecological observations reported within a common set of the 20 papers used for calibration. Variability is expressed as the coefficient of variation (CV; standard deviation divided by mean multiplied by 100) between each observer's overall mean, and as the mean CV of observers' estimates for individual records. }
\vspace{-4 pt}
\begin{center}
\begin{tabular}{rR{1.5cm}R{1.5cm}R{1.5cm}R{1.5cm}R{1.5cm}R{1.5cm}}
\hline
& \multicolumn{3}{c}{\textbf{Spatial}} & \multicolumn{3}{c}{\textbf{Temporal}}\\\cline{2-7}
%Value & Resolution & Extent & \multirow{2}{*}{Actual} & Interval & Duration & Actual Duration \\
Value & Resolution & Extent & Actual Extent & Interval & Duration & Actual Duration \\
\hline
CV of overall mean & 50 & 50 & 75 & 48 & 34 & 109 \\
Mean of record-wise CV & 57 & 41 & 72 & 104 & 64 & 124 \\
\hline
\end{tabular}
\end{center}
\label{default}
\end{table}%

@@

\subsubsection*{Trends in observational methods and scales by year}
\vspace{-10pt}
Figures S1 and S2 respectively show how the frequency of ecological observing methods and the average scale in each dimensions varies by publication year (2004-2014).

@@

\begin{figure}[ht]
%\begin{wrapfigure}{c}{1\textwidth}
\center{\includegraphics[width=0.8\textwidth]{figures/dim_by_yr.png}}
\vspace{5 pt}
\caption{Trends in observational scales by year of publication. The coefficient of a weighted (by number of studies in each year) linear regression, fit to the logarithm (base 10) of the mean scale values (calculated for each publication year) for the six assessed dimensions ispresented at the top of each plot, as well as the model coefficient of determination and p-value. }
\label{afoto1}
\end{figure}
%\clearpage

\subsubsection*{Choice of bandwidth in kernel density estimation}
\vspace{-10pt}
Figure S3 indicates the effect that varying bandwidth has on the appearance of the kernel density estimates. The smallest bandwidth (0.4) tested (Fig. S3 top) shows that the primary observational concentrations described in the main text are evident, but tend to be devolved into separate concentrations. An example is the oblong concentration of observations that in Fig. 1A (and in the lower left panel in Fig. S3) is bounded on the lower right at monthly to yearly intervals and 100-1000 m$^2$ resolutions and on the upper left by near-daily to monthly observations and 0.1-10 ha resolutions. With the smaller bandwidth this concentration appears as two separate patches (Fig. S2 top left), but with 0.7 bandwidth applied becomes coherent (Fig. S3 middle left).  This increasing separation of points into separate clusters is expected as bandwidth is reduced.

\begin{figure}[!ht]
%\begin{wrapfigure}{c}{1\textwidth}
\includegraphics[width=0.9\textwidth]{figures/res_v_extent_ksize2.pdf}
\vspace{10 pt}
\caption{Two-dimensional kernel density estimates of observational densities within the domains defined by sampling interval and spatial resolution (left column) and duration and extent (right column), applied to log-transformed values of each observational dimension. Rows indicate the effects of selecting different bandwidths: 0.4 (top row); 0.7 (middle row); 1 (bottom row). }
\label{afoto1}
\end{figure}

### 2017-05-25 paper/main.tex
Finalized (or nearly) SI, and main text revised up to final section, and final part of abstract.

%Adding to our motivation to record actual observational scales was the fact that 
There were several potential methodological aspects that could have influenced our assessment of ecology's spatial and temporal domains. The first stems from our finding that many studies did not precisely report observational scales, which meant that we had to estimate, rather than simply record, these values for most observations (specifically, in 63, 60, and 67\% of cases for resolution, extent, and actual extent, and 36\%, 64\%, and 83\% of cases for interval, duration, and effective duration). The inevitable errors caused by estimation may have biased our overall findings. However, we attempted to quantify this error by assessing inter-rater disagreement and resampling techniques. The resulting confidence intervals (Fig. 1) suggest that it was unlikely that estimation errors unduly influenced our findings.

Another potential source of bias was the rule set we used to estimate observational scales.

Finally, because our review did not include papers beyond 2014, the omission of studies from the most recent years could have introduced bias into our domain estimates. Indeed, if the trend towards increasing use of remote sensing between 2004-2014 was not spurious, we can project that a repeated study applied to papers published between 2004-2017 would find that remote sensing was used for 7.7\% of observations (a 22\% increase), which would increase the mean effective extent by 17.4\% (95\% CI = -1.3-67\%), or 0.07 orders of magnitude, above that observed here (see SI for details of calculation). Further support for such trend can be found by averaging dimensions by publication year (SI), which reveals that observation extent increased by 0.253 orders of magnitude per year between 2004-2014 (R$^2$ = 0.2.5, p$<$0.07). This somewhat clearer trend also suggests that including more recent studies would show observational extent to be somewhat larger, although in this case by a more modest 5.5\% (0.02 orders of magnitude).

@@

\noindent \textbf{\emph{Insights into the scales of modern ecology}}\\
%These results indicate that for most ecological observations are collected with spatially discontinuous replicates that sample very small portions of the area in which they are distributed.

the continued dominance of field-based research, and the limited use of methods that allow larger areas to be comprehensively observed, such as remote sensing.

satellite observations typically have lower information content than field measurements for a given location, and in many cases only provide proxy measures for the ecological features of interest, such as forest understorey structure \cite{estes_remote_2010}, thereby making ecologists less inclined to use the technology \cite{turner_remote_2003}.

>>>>

%Adding to our motivation to record actual observational scales was the fact that 
There were several potential methodological aspects that could have influenced our assessment of ecology's spatial and temporal domains. The first stems from our finding that many studies did not precisely report observational scales, which meant that we had to estimate, rather than simply record, these values for most observations (specifically, in 63, 60, and 67\% of cases for resolution, extent, and actual extent, and 36\%, 64\%, and 83\% of cases for interval, duration, and effective duration). The inevitable estimation errors may have biased our overall findings. However, we attempted to quantify this error by assessing inter-rater disagreement and resampling techniques. The resulting confidence intervals (Fig. 1) suggest that it was unlikely that estimation errors unduly influenced our findings.

Another potential source of bias was the rule set we used to estimate observational scales, chiefly through our definition of resolution as the smallest area in which all features of interest were measured. We made this choice because we wanted to use a consistent standard. However, some papers reported plot resolutions wherein sub-samples were made within the plot area, thus the average resolution (and actual extent) calculated from our estimates is likely to be smaller than it would be if we had simply recorded reported resolution in all such cases. In addition to this, our decision to exclude experimentally manipulated observations is also likely to have influenced the distributions of all assessed dimensions. For example, the average values of estimated resolutions would likely be smaller if we had included experiments.

Finally, because our review did not include papers beyond 2014, the omission of studies from the most recent years could have introduced bias into our domain estimates. Indeed, if the trend towards increasing use of remote sensing between 2004-2014 was not spurious, we can project that a repeated study applied to papers published between 2004-2017 would find that remote sensing was used for 7.7\% of observations (a 22\% increase), which would increase the mean extent by 17.4\% (95\% CI = -1.3-67\%), or 0.07 orders of magnitude, above that observed here (see SI for details of calculation). Further support for such a trend can be found by averaging dimensions by publication year (SI), which reveals that observational extent increased by 0.253 orders of magnitude per year between 2004-2014 (R$^2$ = 0.2.5, p$<$0.07). This somewhat clearer trend also suggests that including more recent studies would show extent to be somewhat larger, although in this case by a more modest 5.5\% (0.02 orders of magnitude).

@@

\noindent \textbf{\emph{Insights into the scales of modern ecology}}\\
The continued dominance of field-based research, and the limited use of methods that allow larger areas to be comprehensively observed, such as remote sensing.

%These results indicate that for most ecological observations are collected with spatially discontinuous replicates that sample very small portions of the area in which they are distributed.

Satellite observations typically have lower information content than field measurements for a given location, and in many cases only provide proxy measures for the ecological features of interest, such as forest understorey structure \cite{estes_remote_2010}, thereby making ecologists less inclined to use the technology \cite{turner_remote_2003}.
